---
title: Lecture 12 - Data Structures (Part 2)
subtitle: "DSE 511"
author: "Drew Schmidt"
date: "2022-10-04"
output:
  xaringan::moon_reader:
    css: ["metropolis", "includes/theme/custom.css", "includes/theme/2col.css"]
    lib_dir: includes/libs
    nature:
      ratio: "16:9"
      highlightStyle: monokai
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "includes/theme/macros.js"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


# Announcements
* Nothing unresolved from last time
* Homework is live!
    * Due date moved to Wed Oct 5 (midnight)
    * New homework *next* week!
* No class Thursday Oct 6
* Questions?

---
# What's The Goal
* Better code organization
* Engineering over scripting
* Encapsulation over globals

---
# Content
* Lists
* Classes
* Dictionaries



---
class: clear, inverse, middle, center
# Lists

---
# Lists
* An ordered collection of *objects*
* Can store anything
    * Arrays
    * Dataframes
    * Other lists
    * Custom class objects
* Can be implemented in different ways

---
# Lists vs Arrays
.pull-left[
## Lists
Ordered collection of a objects.
].pull-right[
## Arrays
Ordered collection of a values from a single fundamental type.
]

---
# Lists in R and Python
.pull-left[
## R
```{r}
x = list(1, "2", list(a=1, b=2))
x
```
].pull-right[
## Python
```python
x = [1, '2', {'a':1, 'b':2}]
```
```
[1, '2', {'a': 1, 'b': 2}]
```
]

---
# Lists
* Extremely important in R
    * Easy way to handle multi-value returns
    * S3 class decorator/generics
    * ...
* A place to put things
* Helpful for managing lots of things programmatically



---
class: clear, inverse, middle, center
# Classes

---
# Classes and Objects
.pull-left[
* Object Oriented Programming (OOP)
  * Useful organizational strategy
  * Not the only paradigm
* A *class* is a custom data type
    * Describes the data
    * Defines the methods
* It describes an *object*, or instantiation of the class
* The methods operate on the objects
* Methods may use side-effects
].pull-right[
## Pseudocode
```python
x = create_class_object()
x.compute()
y = x.get_result()
```
]

---
# OOP Systems
.pull-left[
## R
* S3
* S4
* R6 (CRAN package)
].pull-right[
## Python
* The `class` system
]

---
# Why Classes?
* Encapsulation
* No ambiguity over function arguments
* Solves the "state problem"

---
# Encapsulation
* Only an object has access to class methods
* Same method name across different classes do different things
* Don't have to worry about unifying different objects under one API
* Useful if you're bad at naming things!

---
# Function Arguments
.pull-left[
## Without OO
```r
x = construct_object()

f = function(x) {
  # ...
}
x = f(x)
```
].pull-right[
## With OO
```r
x = NewObject()
x$f()
```
]

---
# State
* Try not to use global variables
* But what if you have a lot of "moving pieces"
    ```r
    my_fun(x, y, z, w, a, b, c, d, e, ...)
    ```
* Maybe you can re-organize as a class
    * Stuff the data into the object
    * Operate on options internally
    ```r
    x$my_fun()
    ```

---
# Customer Class
.pull-left[
```{r, customer_class, cache=TRUE}
Customer = R6::R6Class(
  public = list(
    initialize = function() {
      self$id = private$create_id()
      self$orders = 0
      self$spend = 0
    },
    increment_orders = function() {
      self$orders = self$orders + 1
    },
    increment_spend = function(spend) {
      self$spend = self$spend + spend
    },
    id = NULL, orders = NULL, spend = NULL
  ),
  private = list(
    create_id = function() {
      httr::GET(url = "https://www.uuidgenerator.net/api/version4") |>
        httr::content()
    }
  )
)
```
].pull-right[
```{r, dependson="customer_class", cache=TRUE}
customer = Customer$new()
customer$id
customer$increment_orders()
customer$orders
new_spend = 12.34
customer$increment_spend(new_spend)
customer$spend
```
]



---
class: clear, inverse, middle, center
# Dictionaries

---
# Dictionaries
* Key/value storage
* Like an in-memory json
* Indexed by keys (not necessarily numbers)
* Major operations
    * Insertion
    * Deletion
    * Retrieval
* Often implemented as a hash table

---
# Hash Table vs Lists
What's the value for the key `"434c8fb1-c329-4234-891d-60bf0a14791e"` ?
.pull-left[
## List
* Check the first element
* Check the second element
* ...
].pull-right[
## Hash Table
* Run key through "hash function"
* Get value at index output by hash function
]

---
# Hash Table vs Lists
What's the value for the key `"434c8fb1-c329-4234-891d-60bf0a14791e"` ?
.pull-left[
## List
Runtime scales with the number of elements in the list
].pull-right[
## Hash Table
Runtime is fixed regardless of the size of the hash table
]

---
# Dictionaries in Python
```python
d = {}
d['key1'] = 'val1'
d['key2'] = 'val2'
d
```
```
{'key1': 'val1', 'key2': 'val2'}
```
```python
d['key1']
d['key1'] = 'newval'
d
```
```
{'key1': 'newval', 'key2': 'val2'}
```

---
# Building a Dictionary in R
* R has no "dictionary" built in
* It does have hash tables (environments)
* We can implement a dictionary using a hash table...

---
# Dictionary
```{r, dict, echo=FALSE, cache=TRUE}
Dictionary = R6::R6Class(
  public = list(
    initialize = function() {
      private$env = new.env()
      invisible(self)
    },
    length = function() {
      base::length(private$env)
    },
    print = function() {
      cat("A dictionary with", self$length(), "elements\n\n")
    },
    insert = function(key, value) {
      private$env[[key]] = value
      invisible(self)
    },
    update = function(dict) {
      keys_new = dict$keys()
      for (key in keys_new)
        self$insert(key, dict$get(key))
    },
    get = function(key) {
      private$env[[key]]
    },
    clear = function() {
      private$env = NULL
      invisible(gc())
      private$env = new.env()
    },
    keys = function() {
      ls(private$env)
    },
    values = function() {
      as.list(private$env)
    }
  ),
  private = list(
    env = NULL
  )
)
```

.pull-left[
```r
Dictionary = R6::R6Class(
  public = list(
    initialize = function() {
      private$env = new.env()
      invisible(self)
    },
    length = function() {
      base::length(private$env)
    },
    print = function() {
      cat("## A dictionary with", self$length(), "elements\n\n")
    }
  ),
  private = list(
    env = NULL
  )
)
```
].pull-right[
```{r, dependson="dict", cache=TRUE}
d = Dictionary$new()
d
```
]

---
# Dictionary
.pull-left[
```r
    # ...
    insert = function(key, value) {
      private$env[[key]] = value
      invisible(self)
    },
    get = function(key) {
      private$env[[key]]
    },
    # ...
```
].pull-right[
```{r, dependson="dict", cache=TRUE}
d = Dictionary$new()
d
d$insert("x", 1)
d$get("x")
d
```
]

---
# Dictionary
.pull-left[
```r
    # ..
    keys = function() {
      ls(private$env)
    },
    values = function() {
      as.list(private$env)
    }
    # ..
```
].pull-right[
```{r, dependson="dict", cache=TRUE}
d = Dictionary$new()
d$insert("x", 1)
d$values()
```
]

---
# Dictionary
.pull-left[
```r
    # ...
    update = function(dict) {
      keys_new = dict$keys()
      for (key in keys_new)
        self$insert(key, dict$get(key))
    },
    # ...
```
].pull-right[
```{r, dependson="dict", cache=TRUE}
d = Dictionary$new()
d$insert("x", 1)

new_dict = Dictionary$new()
new_dict$insert("y", 2)
d$update(new_dict)

d
d$values()
```
]

---
# Dictionary
.pull-left[
```r
    # ...
    clear = function() {
      rm(private$env)
      invisible(gc())
      private$env = new.env()
    },
    # ...
```
].pull-right[
```{r, dependson="dict", cache=TRUE}
d = Dictionary$new()
d$insert("x", 1)
d
d$clear()
d
d$values()
```
]

---
# A Dictionary of Customers
```{r, customers_dict, dependson=c("dict", "customer_class"), cache=TRUE}
create_customer = function(customers) {
  customer = Customer$new()
  customers$insert(customer$id, customer)
  customers
}

customers = Dictionary$new()
for (i in 1:5) {
  customers = create_customer(customers)
}

keys = customers$keys()
keys
```

---
# A Dictionary of Customers
```{r, dependson=c("dict", "customer_class", "customers_dict"), cache=TRUE}
customer_id = keys[1]
customer = customers$get(customer_id)

customer$increment_orders()
customer$orders

new_spend = 7.89
customer$increment_spend(new_spend)
customer$spend

customers$insert(customer_id, customer)
customers$get(customer_id)$spend
```



---
class: clear, inverse, middle, center
# Wrapup

---
# Wrapup
* Classes
    * Can be a helpful way of organizing code
    * Not a "suicide pact"
* Lists and dictionaries are useful data structures

---
class: clear, inverse, middle, center
# Questions?
